---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
Use the data.table to import and query against the dataset
```{r cache=F}

library(data.table)

library(lubridate)

act <- read.table(file="activity.csv", header=T, sep=",", row.names=NULL, stringsAsFactors=F)

## convert the date to actual date type
act$date <- ymd(act$date)

attach(act)

```

## What is mean total number of steps taken per day?
```{r}
library(dplyr)

stepsPerDay <- act %>% filter(complete.cases(act)) %>% select(date, steps) %>% group_by(date) %>% summarize(dailyCount = sum(steps))
print("Confirm sum of data")
head(stepsPerDay)

par(mfrow=c(1,1))
hist(stepsPerDay$dailyCount, breaks=35, xlab="Steps by Day (Should label with date)", ylab="Sum of steps per day", main="Frequency of Steps Per Day")

```

Mean of steps taken per day: 
``` {r} 
print(mean(stepsPerDay$dailyCount))
```

Media of steps taken per day:
``` {r} 
print(median(stepsPerDay$dailyCount))
```


## What is the average daily activity pattern?

```{r}
averStepsPerInterval <- act %>% filter(complete.cases(act)) %>% select(interval, steps) %>% group_by(interval) %>% summarize(avgSteps = mean(steps))

print("Confirm group and avg of data")
head(averStepsPerInterval)

plot(x=averStepsPerInterval$interval, y=averStepsPerInterval$avgSteps, type="l", xlab="Intervals (24 hour on 5 minute increments)", ylab="Average Steps Taken Per Interval", main="Average of Time Intervals Across All Days")

```

Which 5 minute increment contains the maximum number of steps:
```{r}
averStepsPerInterval[max(averStepsPerInterval$avgSteps), ]

``` 
Roughly around 5PM at night.  This probably correlates with leaving work and coming home.


## Imputing missing values
```{r}
misValues <- act %>% filter(!complete.cases(act))
head(misValues)
summary(misValues)
length(misValues[is.na(misValues$steps),1])
```
Total number of missing steps is `r length(misValues[is.na(misValues$steps),1])`

Replace NA values with the average for the given interval period

```{r}

newStepCount <- merge(act, averStepsPerInterval, by="interval")
newStepCount <- newStepCount %>% mutate(steps= ifelse(is.na(steps), avgSteps, steps))

#Confirm that all NA's have been removed.
sum(!complete.cases(newStepCount))

newStepsPerDay <- newStepCount %>% select(date, steps) %>% group_by(date) %>% summarize(dailyCount = sum(steps))
print("Confirm sum of data")
head(newStepsPerDay)
summary(newStepCount)

par(mfrow=c(2,1))
hist(newStepsPerDay$dailyCount, breaks=35, xlab="Steps by Day (Should label with date)", ylab="Sum of steps per day", main="Frequency of Steps Per Day (Modified to remove NA)")

hist(stepsPerDay$dailyCount, breaks=35, xlab="Steps by Day (Should label with date)", ylab="Sum of steps per day", main="Frequency of Steps Per Day (Original Unmodified Data")

```

Mean of steps taken per day (NA replaced with interval mean): 
``` {r} 
print(mean(newStepsPerDay$dailyCount))
```

Media of steps taken per day (NA replaced with interval mean):
``` {r} 
print(median(newStepsPerDay$dailyCount))
```

```

The mean and median values between the original data and the modified data are about the same.  If you look at the actual histograms you can see a bit of shifting.  Some values are higher and some values are lower.  This did not make much of a difference in the mean values even though we can seem some shifting in the histogram.

## Are there differences in activity patterns between weekdays and weekends?
